---
title: "Gene Set Enrichment Analysis"
output:
  html_document:
    theme: flatly
    df_print: paged
    css: "markdown-styles.css"
    toc: true
    toc_depth: 3

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE, class.source = "code-chunk", out.width = "100%", fig.show = "hold")

devtools::load_all()
library(tidyverse)

object_t269 <- loadSpataobject_t269("data/object_t269_UKF269T.RDS")

```

## 1. Prerequisites 

- [Differential Expression Analysis](dea.html)
- [Molecular Signatures](molecular-signatures.html)

## 2. Introduction

For a deeper into the results of the DEA one can look for gene sets that are enriched in the proposed clustering or created based on histology. `SPATA2` implements the [`hypeR`](https://github.com/montilab/hypeR) package which uses hypergeometric testing for enriched gene sets.

```{r eval = F}

library(SPATA2)
library(SPATAData)
library(tidyverse)

# download sample UKF269T
object_t269 <- downloadFromPublication(pub = "kueckelhaus_et_al_2024", id = "UKF269T")

# add the grouping variables if you created object_t269 yourself from raw data 
data("spatial_segmentations")

object_t269 <- 
  addFeatures(
    object_t269 = object_t269, 
    feature_df = spatial_segmentations$UKF269T, 
    overwrite = TRUE
  )

# histology only
plotSurface(object_t269 = object_t269, pt_alpha = 0)

# colored by clustering
plotSurface(object_t269 = object_t269, color_by = "histology")

```

```{r echo = F, out.width = "50%", fig.cap = "Fig.1 Shared Nearest Neighbour clustering will be analyzed with GSEA."}

# histology only
plotSurface(object_t269 = object_t269, pt_alpha = 0)

# colored by clustering
plotSurface(object_t269 = object_t269, color_by = "histology")

```

## 3. Running Gene Set Enrichment Analysis (GSEA)

The function `runGSEA()` conducts the computation. It requires the results from `runDEA()` which conducts the differential expressiona nalysis. By default, the function uses all gene sets that are saved in the `SPATA2` object_t269. If there are gene sets that you don’t want to test against you can provide a subsetted list of gene sets. The default `SPATA2` object_t269 contains a variety of gene sets of multiple classes (Hallmark, Biocarta, etc.) marked with a corresponding prefix like *HM* or *BC*. 

```{r}

gs_list <- getGeneSetList(object_t269)

length(gs_list)

```

For the sake of clarity, this example only uses Biocarta (BC) and Hallmark (HM) gene sets.

```{r}

# subset gene sets that start with HM or BC
sub_vec <- str_detect(names(gs_list), pattern = "^HM|^BC")

gs_list_sub <- gs_list[sub_vec]

length(gs_list_sub)

set.seed(1)
sample(gs_list_sub, size = 3)

```

Use argument `signatures` if you want to run GSEA only on a subset of gene sets. Keep it empty if you want to test against all gene sets stored in the `SPATA2` object_t269. This example only uses Hallmark gene sets. 

```{r}

object_t269 <- 
  runGSEA(
    object_t269 = object_t269,
    across = "histology",
    signatures = names(gs_list_sub)
    )

```

## 4. Extracting results

The results can be manually extracted via the following functions. `getGseaResults()` extracts a list of `hypeR` object_t269s - on for each group. `getGseaResultsDf()` extracts a data.frame that results from merging the data of all `hypeR` object_t269s together. The group belonging of each gene set is saved in the variable/column that is named according to the grouping variable - here *histology*.

```{r}

getGseaDf(
  object_t269 = object_t269, 
  across = "histology",
  method_de = "wilcox", 
  n_gsets = 3 # extract top 20 most significant gene sets
) 

```


## 5. Plotting results
Gene set enrichment results can be visualized via dot plots. This can be either done by group with `by_group = TRUE` or merged for all groups. Figure 2 visualizes the enrichment for cluster 0 and 5 highlighting the hypoxia associated activity in this area.

```{r fig.dim = c(9, 9), fig.cap = "Fig.2 Dot plot to display GSEA results by group."}

plotGseaDotPlot(
  object_t269 = object_t269,
  across = "histology",
  across_subset = c("tumor", "infiltrated"),
  n_gsets = 10,
  by_group = TRUE, 
  transform_with = list(fdr = log10),
  nrow = 2
)

```

Using `by_group = FALSE`, the results are merged to one plot. 

```{r fig.cap = "Fig.3 Merged GSEA dot plot."}

plotGseaDotPlot(
  object_t269 = object_t269,
  across = "histology",
  color_by = "histology",
  n_gsets = 7,
  pt_alpha = 0.8,
  transform_with = list(fdr = log10),
  by_group = FALSE # merge in one plot
) 

```

